On the uniqueness and stability of dictionaries \\ for sparse representation of noisy signals

### 2020-08-04 manuscripts/IEEE/revisions/response.txt

REVIEWER 1

Summary 
The paper provides conditions under which dictionary learning is stable (in particular to noise), which apply in almost all cases. This, the authors claim, provides some support for the empirical observation that different dictionary learning methods lead to very similar dictionaries.

Comments 
The paper is quite well-written. I am not an expert in dictionary learning per se, but if these are truly the first perturbation bounds for dictionary learning, then they are worth publishing in TSP.

Other comments
1. “then each submatrix formed from 2k of its columns or less has a strictly positive lower bound.” What is k here?

The full sentence is: "By compactness of the unit sphere, injective linear maps have nonzero lower bound; hence, if M satisfies (2), then each submatrix formed from 2k of its columns or less has a strictly positive lower bound." Here, k is the sparsity referred to in (2), which states that a matrix M satisfies the spark condition if Mx1 = Mx2 implies x1 = x2, for all k-sparse x1, x2.

@@

2. 3 lines below (5): why is ∪H = [m] required?

The requirement becomes clearer recalling the definition L_k := L_(m choose k). Our statement is then:

@@

3. Are J and Jbar in Theorem 1 of same size?

Yes. We have clarified this ambiguity.

4. About Theorem 1. I am a little worried about C1 depending on A and the xi’s. It could make some of parts of the statement void. Presumably, one sees these as fixed, and looks at the possible B’s and possible \bar{x}_i’s satisfying |Ax_i − B\bar{x}_i|_2 \leq \epsilon?

That is correct. We see no need to worry. Fixing A and the x_i's fixes the numbers L_2k(A) and C_1(A,x_i). Theorem 1 then states that, provided \epsilon < L_2k/C_1, all such B's and \bar{x}_i's are close to A and the x_i's, respectively (see (6) and (7)). To rephrase via the triangle inequality, all good enough B’s and \bar{x}_i’s are similar to one another, or identical when \epsilon = 0.

@@

5. Also, in (7), using xi to denote the restriction of xi to J, and similarly for xi, could lead to confusion, in particular if the reader glances at the result without reading the last sentence.

We have altered the notation to abate this potential confusion.

6. “A practical implication of Thm. 1 is the following: there is an effective procedure sufficient to affirm if a proposed solution to Prob. 1 is indeed unique (up to noise and inherent ambiguities). One need simply to check that the matrix and codes satisfy the (computable) assumptions of Thm. 1 on A and the xi.” I question whether the assumptions of Theorem 1 are indeed computable. Could the authors elaborate on that?

The assumptions of Theorem 1 are:
i) L_2H(A) > 0 for an r-regular H with the SIP.
ii) For each S in H, there are more than (k-1)(mCk) k-sparse vectors x_i supported on S.
iii) \epsilon < L_2(A)/C_1 to guarantee (6) and \epsilon < L_2k(A)/C_1 to guarantee (7)

@@

Conditions (i) and (ii) are obviously computable, though it is another question, and what we assume to be the reviewer's concern, whether or not there exists an efficient algorithm to do so. Condition (iii) relies on the definition of C1, which appears in Section III. Some comments on this appear in section III:

"The pragmatic reader should note that the explicit constants C1 and C2 are effectively computable: the quantity Lk may be calculated as the smallest singular value of a certain matrix, while the quantity ξ involves computing “canonical angles” between subspaces, which reduce again to an efficient singular value decomposition. There is no known fast computation of L in general, however, since even L > 0 is NP-hard [19]; although fixing k yields polynomial complexity. Moreover, calculating C2 requires an exponential number of queries to ξ unless r is held fixed, too (e.g., the “cyclic order” hypergraphs described above have r = k). Thus, as presented, C1 and C2 are not efficiently computable."

7. I am not sure what “substitutions” means in Theorem 3. Does that mean “choices” of A and x1, . . . , xN such that Axi = yi? Isn’t Theorem 3 a simple corollary of Theorem 1, together with the remark in the couple of paragraph above it? If so, I suggest making it a corollary. In the same vein, I am not sure that Proposition 1 is rigorous enough to make it a proposition. It could be stated as an informal claim instead.

"Substitution" here mean such assignments or choices of real numbers to the indeterminate variables, yes. We frame it as a theorem and not a corollary because we draw the important Corollary 2 directly from it. We have no qualms downgrading Proposition 1 to an informal claim, however.

[NOTE: should we just make everything a corollary of Thm 1..? Even Thm 2?]

8. The perturbation bound is used by the authors to provide some explanation for why, in practice, different methods lead to similar dictionaries. Can that also be used to explain the second (bold) part of their observation that “[these waveforms] appear in dictionaries learned by a variety of algorithms trained with respect to different natural image datasets.”? Just curious.

To be clear, this (bold) observation is just an observation which motivates the work. Imagining natural image patches to form a population which satisfies (1) with respect to some ground truth dictionary A, we have shown that there exists a sufficient number of samples to uniquely identify it or something close to it via approximate solutions. Given this fundamental stability and uniqueness of the problem solutions, the fact that a variety of methods geared to approximately solve Problems 1 and 2 all seem to capture similar structure is much less surprising.

@@

REVIEWER 3

Comments:
This paper considers the uniqueness and stability of sparse dictionary learning problem. The paper is well written and organized. Although it is a pure theory paper, it is fairly easy to follow the key ideas without messing up with the proofs. Below are some more detailed comments:

1. This work considers the uniqueness and stability of SIP condition of the dictionary A. One of the fundamental questions is can the Problem 2 be solved to global optimality under these conditions? Because dictionary learning is a bilinear and hence nonconvex problems, solving Problem 2 with l0 norm could still be intractable even though uniqueness condition exists. Can the author comment on this issue of achieving global optimality for solving Problem 2?

Theorem 2 guarantees that all globally optimal solutions to Problem 2 are identical up to noise and inherent permutation/scaling ambiguities. This doesn't change the fact that computing any one such global solution given data is still NP-hard.

@@

As l0 norm is intractable, does all the properties still hold for l1 norm minimization? Can we solve it efficiently (polynomial time) to global optimality? Similar issue for other works, such as
https://arxiv.org/pdf/1807.05595.pdf

We make no claims in this paper about l1-norm minimization, though we are hopeful a clever reader may find some way to build off our results and derive related guarantees for this continuous relaxation of the sparse coding problem.

@@

But heuristic methods are proposed there to check global optimality. A discussion and elaboration on this could be very helpful here.

Yes, we only claim that our general uniqueness criterion provide a (not necessarily efficient) means of checking the globality of a proposed solution. It could be that in certain restricted cases these conditions are efficiently computable, but this is not the case in general.

@@

2. The number of samples N >= (k-1) ( m choose k) for the uniqueness condition grows exponentially with respect to k. Is this condition optimal w.r.t. SIP property, or can be reduced? On the other hand, is there any relationship between SIP and RIP conditions? The lower bound of (5) looks similar to RIP conditions.

We comment on sample complexity in the last paragraph of Section II:

"We close this section with comments on optimality. Our linear scaling for ε in (8) is essentially optimal (e.g., see [23]), but a basic open problem remains: how many samples are necessary to determine the sparse coding model? If k is held fixed or if the size of the support set of reconstructing codes is known to be polynomial in m and k, then a practical (polynomial) amount of data suffices. [footnote: In the latter case, a reexamination of the pigeonholing argument in the proof of Thm. 1 requires a polynomial number of samples distributed over a polynomial number of supports.] Reasons to be skeptical that this holds in general, however, can be found in [19], [24]"

In general, our conditions are sufficient but not proven necessary, and we doubt we have achieved the tightest possible bounds all around.

@@

As for the SIP and RIP, a footnote points out the relation between our matrix lower bound L_H(M) and the RIP:

"We note that $1 - \sqrt{k} L_k(\mathbf{M})$ is known as the asymmetric lower restricted isometry constant for matrices $\mathbf{M}$ with unit $\ell_2$-norm columns"

@@

3. Corollary 2 suggests that A and x satisfying SIP can be generated by drawing the entries independently from a continuous distribution. This is often unrealizable in practice. Is there any deterministic way of generating A and x with SIP condition?

In the prose surrounding Corollary 1 we give a deterministic construction of sparse vectors x_i satisfying the constraints of Theorem 1.

TODO : can we deterministically generate matrices injective on SIP hyper graphs? The (mCk)-hypergraph is one such possibility, so maybe not in general? But for special cases (convolution matrices..?)

### 2020-08-04 manuscripts/PNAS/first review/point_by_point_response.txt

-----------------
Since the result is an extension of the noise-free result of (HS15) to a noisy case, the authors should provide better justification for the significance of contributions. The main result seems to be based on an approach similar to (HS15), and there is little to no discussion of the computability and tightness of the sufficient condition for stability in the main results. Thus, the submission does not seem to be greatly more valuable than (HS15) and also not conclusive regarding the central problems of sparse linear coding and dictionary learning. That being said, the paper (with the modifications below) would certainly merit publication in very solid journal like IEEE Transactions on Information Theory, where it would be a median paper, in my opinion.

-----------------

RESPONSE:

We agree with the reviewer that a straightforward extension of the approach and noise-free result of (HS15) to the noisy case — while still a significant advancement in applied mathematics with actual practical implications — would not necessarily merit publication in PNAS on its own. However, our main result (Thm. 1) generalizes the noise-free result of (HS15) far beyond a simple extension to the noisy case in ways that provide valuable insight into the dictionary learning model; and, moreover, a significant deviation from the approach taken in (HS15) was necessary to prove these new facts. Additionally, our solution (Cor. 2) to the optimization formulation of the dictionary learning problem (Prob. 2) — of most interest to optimization theorists and practitioners of their methods — is (to our knowledge) the first of its kind in both the noise-free and noisy domains. We discuss all of these points in more detail in our response to the reviewer’s later "detailed comments".

@@

As per the reviewer’s request, we have included in our updated manuscript a discussion of the computability of the sufficient conditions (they are, indeed, computable). The main constants in our work -- C_1, C_2 -- are computable, elementarily, in terms of two positive quantities that are natural in our context: (a) the restricted "Lower Bound" L_k(M) of a matrix M, which is a measure of how distinguishable two (k/2)-sparse x’s are by their images under M, and (b) the value in Def. 4 of \xi(V_1,..,V_l) given subspaces V_i spanned by columns of A, which bounds the distance between intersections of such subspaces and each subspace individually. The first quantity (a) is, in some form or another, related to standard objects in the field and may be computed as the smallest singular value of a set of specific sub-matrices; in particular, for fixed sparsity constraint k, the lower bound L_k(A) is efficiently computable (polynomial time). Because deciding L_k(A) > 0 is already NP-hard (TP14), it seems likely that without fixing a sparsity level there is no efficient computation in general for (a). The calculation of (b) involves determining "canonical angles" between subspaces, which is usually reduced again to a numerical (efficient) singular value decomposition given the fixed subspaces. However, upon close inspection, C_2 nonetheless requires an exponential number of instances of computing \xi. Thus, while our constants are effective (an algorithm exists to compute them), they are not necessarily polynomial-time computable.

We next respond to the reviewer's question about the mathematical "tightness" for the three types of quantities in our results : (a) required coding error eps(delta_1, delta_2) sufficient to guarantee uniqueness in terms of desired dictionary, delta_1, and sparse representation error, delta_2; (b) the constants C_1 and C_2 involved in our sufficient linear bound on eps(delta_1, delta_2) in terms of the noise; and (c) the number of sufficient samples N guaranteeing uniqueness.  With respect to (a), even with fixed dictionary, the recovery of sparse representations up to an error that is linear in the noise is essentially optimal (ACD13) (up to constants). Thus, the fact that we have a sufficient error in coding for uniqueness being (asymptotically) linear for both the desired dictionary and sparse vector recovery error is optimal.  Although optimal, our solution to Problem 1 involves constants C_1 and C_2, and it would be interesting and relevant for practical applications to determine the minimal such constants. In the case of sparsity k=1, we have given an example showing optimality of constants in this case. Finally, with respect to (c), we regard this as a fundamental open problem in the field: how many sparsely generated samples determine the linear coding model?  (One reason to be skeptical that a polynomial suffice is (TP14, T15).)  However, there is some good news -- if k is fixed or if the sparse codes \bar x_i are known to come from a polynomial number of supports, then a polynomial number of samples N is enough to guarantee uniqueness (which, incidentally, does not follow from the results of (HS15) in the noiseless case). We have revised the manuscript to discuss these points.

@@

-- The paper does a good job of motivating the general problem of sparse coding to a general audience. In particular, there is a good review of the areas in which dictionary learning arises.

-- I think the biggest potential issue for publication in PNAS is that the significance of the result needs more discussion. Specifically, the authors could address why going from the noise-free case to the noisy case is not incremental. PNAS specifically targets submissions that go beyond what could be published in a more specialized journal. Could the authors, for example, contrast the proof technique between the noisy and noise-free cases? Or, perhaps, the practical implications?
-----------------

RESPONSE:

With all due respect to the reviewer, we believe the most compelling practical implication of a noisy result over a noiseless result is plainly evident: never has there been an experiment without noise and an unstable model is therefore useless in practice. We assume the reviewer must instead be concerned with what additional practical "take-away" message our manuscript contains.  One significant one is the outline of a procedure that can (sufficiently) affirm if one’s proposed solution to Prob. 1 or Prob. 2 is indeed unique. We have updated the manuscript with an explicit statement of this fact, and we outline one such procedure here:

Given a dictionary A and codes x_i that solve Problem 1 (e.g., learned by any algorithm), check that that they satisfy the assumptions of Thm. 1:
   - List the support sets of the x_i.
   - Discard those for which there are (k-1){m \choose k} or fewer x_i with that support.
   - Discard those for which the supported x_i are not in general linear position. (This should also exclude any supports less than k in size.)
   - From the support sets that remain, list all SIP-satisfying hypergraphs that are subsets of this set.
   - Discard those for which L_{2H}(A) = 0.
   - For each of the remaining hypergraphs, determine the constant C_1 from A and the x_i (or for every subset of the x_i of size (k-1){m \choose k}).
   - Check that the derived upper bound on \varepsilon exceeds that of the original problem. If yes, the solution is "unique".

@@

There are other practical (polynomial time) implications of our discovery of sufficient combinatorial designs for support sets of generating codes x_i. Moreover, we have shown that a subset of dictionary elements are recoverable even if the number of dictionary elements in total is unknown; these observations are discussed in more detail below.

The reviewer also raises the concern that, regardless of practical implications, our results may amount to an incremental advance over those of (HS15) by way of similar techniques. We disagree with this assessment on both counts: our main result (Thm. 1) goes far beyond a straightforward extension of that in (HS15) to the noisy case and this required a significant deviation from their approach.

It is understandable that the reviewer may have thought otherwise, considering how the proof of Thm. 1 is presented in the manuscript. As observed by our second reviewer, however, the extension to the noisy case did indeed require a novel combination of several results in the literature. Specifically, the main difficulty was to generalize Lemma 1 to the case where the k-dimensional subspaces spanned by corresponding (through the map \pi) submatrices of A and B are assumed only to be proximal (small ‘d’), and not identical as in (HS15). In contrast to the noiseless case, here it must be explicitly demonstrated that this proximity relation is propagated through the repeated intersections of these submatrix spans all the way down to the spans of dictionary elements themselves. We designed and proved Lemma 3 to address this issue, which draws its bound from the convergence guarantees of an alternating projections algorithm first proposed by von Neumann. This result, combined with a little known fact about the distance ‘d’ requiring proof in (M10), constitute the more obscure components of the deduction in Eq. (26). To reiterate, this step is completely trivial in the noiseless case and required no mention for the inductive steps taken in (HS15).

Our proof of Lemma 1 diverges perhaps even more significantly from the approach taken in (HS15) by way of Lemma 4. Key to our reduction of the sample complexity given in (HS15) by an exponential factor is the introduction of a combinatorial design (the "singleton intersection property") for support sets. Since in this case the map \pi from supports in the hypergraph to {[m] \choose k} is not surjective, one can not apply the same inductive method as in (HS15), which freely chooses supports in the codomain to intersect at (k-1) nodes and map back to some corresponding (k-1)-sized intersection of supports in the domain. Instead, we demonstrate the surprising fact that by pigeonholing the images of supports in the (SIP-satisfying) hypergraph H, one can still guarantee a bijection between the nodes and therefore the subspaces spanned by individual dictionary elements. We note that it was necessary to forgo inductive methods altogether in order to prove this fact for all hypergraphs satisfying the SIP; otherwise, we would require that the supports in the hypergraph have intersections of size k’ for every k’ < k (e.g., this is not the case for the small SIP example we give consisting of the rows and columns of nodes arranged in a square grid). Again, to our surprise, it so happens that this new induction-less argument easily generalizes to the case where B has an arbitrary number of columns, in which case we find that a one-to-one correspondence exists between a subset of columns of A and B of a size that has a nice closed-form expression for regular SIP hypergraphs.

@@

To be clear, the new perspective we take on the problem yields the following powerful conclusions beyond those of a straightforward extension of (HS15) to the noisy case:

1) An extension to the case where the number of dictionary elements is unknown: The results of (HS15) only apply to the case where the matrix B has the same number of columns as A. We forgo this assumption and show that B must have at least as many columns as A and contains (up to noise) a subset of the columns of A. The size of this subset depends on a simple relation between the number of columns of B and the regularity of the support-set hypergraph.

2) A significant reduction in sample complexity: To identify the n x m generating dictionary A, (HS15) require that data be sampled from every k-dimensional subspace spanned by the m columns of A (that is, {m \choose k} subspaces in total). We show that the data need only be sampled from m subspaces in total (e.g. those supported by consecutive intervals of length k in some cyclic order on [m]) and in some cases as few as 2\sqrt{m} (e.g. when m = k^2, take the supports to be the rows and columns of [m] arranged in a square grid). Moreover, if the size of the support set of reconstructing codes is known to be polynomial in m and k, then the pigeonholing argument in the proof of Thm. 1 requires only a polynomial number of samples distributed over a polynomial number of supports; thus, N is then polynomial in \bar m, k. This point was only hinted at in the Discussion section of our original submission, but we have included it in the updated manuscript to make clear the power of our approach over that taken in (HS15).

3) No spark condition requirement for the generating matrix A: One of the mathematically significant contributions made by (HS15) was to forgo the constraint that the recovery matrix B also satisfy the spark condition, in contrast to all previously known uniqueness results which (either explicitly or implicitly) made this assumption.  Our proof is powerful enough to show that, in fact, even the matrix A need not satisfy the spark condition to be identifiable! Rather, it need only be injective on the union of subspaces with supports that form sets in a hypergraph satisfying the singleton intersection property (SIP). (For example, consider the matrix A = [e_1,.., e_5, v] where v = e_1 + e_3 + e_5, and take H to be the set of all consecutive pairs of [m] arranged in cyclic order. Then A satisfies the assumptions of Thm. 1 for dictionary recovery without satisfying the spark condition.) We had omitted this point in our original submission of the manuscript to keep things simple, but we have decided to include this fact in our revision. This also required us to redefine the restricted matrix lower bound L_k to be in terms of a hypergraph H (L_H in the revised manuscript), which is an interesting object for further study in its own right.

@@

The work presented is analytical and does not rely on data validation.
-----------------

RESPONSE: Both reviewers have raised this concern and we again thank the editor for the opportunity to address it. It seems both reviewers have not noticed the several ways in which our results generalize the modern results of (HS15) beyond a straightforward extension to the noisy case, not to mention our solution to the optimization Prob. 2, which as far as we know is the first of its kind and of most interest to practitioners. We have already discussed these points in detail in our response to the first reviewer’s identical comment.

-----------------
In the paragraph "Since f is an analytic function..." (page 3) it is argued that a specific construction of a suitable matrix implies that "almost every real nxm matrix... satisfies Eq. (2)". This appears to overstate the result, in my opinion - can the authors elaborate on why "almost every real matrix" can be obtained from this construction based on a Vandermonde matrix with additional zero rows?
-----------------

RESPONSE: The point we make here is not that almost every real matrix satisfying Eq. (2) can be obtained from the construction based on a Vandermonde matrix which we provide; rather, the point is that almost every matrix satisfies Eq. (2) because at least one matrix does (e.g., our construction) and therefore, since f is analytic, the set of all those that do not satisfy the condition has measure zero. We have updated the manuscript to clarify this argument for our readers.

### 2020-08-04 manuscripts/arxiv/final_submission.tex
everything

% Notes:
%
% Can we prove in a line or two that the inverse problem is stable by arguing that the function is bijective and closed therefore a homeomorphism?
% Should we mention the notion of Grassmannian manifolds and that Theta is a metric on this space?
% Change robustness to stability?
%
\documentclass[journal, twocolumn]{IEEEtran}

@@

\begin{figure}
\begin{center}
\includegraphics[width=.24 \linewidth]{manuscripts/IEEE/figs/Basis_.png}
% \includegraphics[width=.25 \linewidth]{figs/Ahat_FastICA_M640_K8_noise0_2042}
%\includegraphics[width=.24 \linewidth]{figs/Ahat_FastICA_M640_K8_noise0_4083.png}
\includegraphics[width=.24 \linewidth]{manuscripts/IEEE/figs/Ahat_FastICA_M1280_K8_noise0_0001.png}
\includegraphics[width=.24 \linewidth]{manuscripts/IEEE/figs/Ahat_FastICA_M1280_K8_noise2_0513.png}
\includegraphics[width=.24 \linewidth]{manuscripts/IEEE/figs/Ahat_FastICA_M1280_K8_noise3_9744.png}
\caption{\textbf{Learning a dictionary from increasingly noisy data}. The (unraveled) basis elements of the $8 \times 8$ discrete cosine transform (DCT) form the 64 columns of the left-most matrix above. Three increasingly imprecise dictionaries (columns reordered to best match original) are recovered by FastICA \cite{hyvarinen2000independent} trained on data generated from $8$-sparse linear combinations of DCT elements corrupted with additive noise (increasing from left to right).}
\vspace{-.6 cm}
\label{noisyrecovery}
\end{center}
\end{figure}

>>>>

% Notes:
%
% Can we prove in a line or two that the inverse problem is stable by arguing that the function is bijective and closed therefore a homeomorphism?
% Change robustness to stability?
%
\documentclass[journal, twocolumn]{IEEEtran}

@@

\begin{figure}
\begin{center}
\includegraphics[width=.24 \linewidth]{Basis.png}
\includegraphics[width=.24 \linewidth]{Ahat_FastICA_M640_K8_noise1_0205.png}
\includegraphics[width=.24 \linewidth]{Ahat_FastICA_M640_K8_noise2_0409.png}
\includegraphics[width=.24 \linewidth]{Ahat_FastICA_M640_K8_noise5_1021.png}
\caption{\textbf{Learning a dictionary from increasingly noisy data}. The (unraveled) basis elements of the $8 \times 8$ discrete cosine transform (DCT) form the 64 columns of the left-most matrix above. Three increasingly imprecise dictionaries (columns reordered to best match original) are recovered by FastICA \cite{hyvarinen2000independent} trained on data generated from $8$-sparse linear combinations of DCT elements corrupted with additive noise (increasing from left to right).}
\vspace{-.6 cm}
\label{noisyrecovery}
\end{center}
\end{figure}
